---
title: "Data Processing Steps"
author: "William May"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Processing Steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
urlcolor: blue
---

```{r setup, include=FALSE}
library(knitr)
library(graph)
library(Rgraphviz)
opts_chunk$set(echo = FALSE)
```

```{css}
.caption {
	font-style: italic;
}
```

Data processing instructions are stored as a set of scripts under the `analysis`
folder. The following flowchart shows the relationship between scripts and data
files, leading to the final packaged dataset:

```{r scripts, fig.height = 8.5, fig.cap='Processing flowchart. Green boxes are data files and gray ellipses are scripts. Text in angle brackets depends on options provided to the script.'}
scripts = c('clean_old_routine', 'load_raw', 'load_calibration',
            'process_new_data', 'aggregate_hourly', 'routine_package')
scripts = paste0(scripts, '.R')
files = c('old hourly files', 'measurement files', 'calibraton files',
          'old_<site>.csv', 'raw_<site>_<logger>.sqlite', 'cals_<site>.sqlite',
          'processed_<site>_<logger>.sqlite', 'hourly_<site>_<logger>.sqlite',
          'routine_chemistry_v<version>.zip')
nodes = c(scripts, files)
nnodes = length(nodes)
gadj = matrix(0, nnodes, nnodes, dimnames = list(nodes, nodes))
edge_inds = c(
    7, 1,
    1, 10,
    8, 2,
    2, 11,
    9, 3,
    11, 3,
    3, 12,
    11, 4,
    12, 4,
    4, 13,
    13, 5,
    5, 14,
    10, 6,
    14, 6,
    6, 15
)
gadj[matrix(edge_inds, ncol = 2, byrow = T)] = 1
g = graphAM(gadj, 'directed')

attrs = list(node = list(shape = 'ellipse', fixedsize = F, fillcolor = 'gray88',
                         fontsize = 15))
nAttrs = list(shape = sapply(files, function(x) 'box'),
              fillcolor = sapply(files, function(x) '#ADF0BB'))
plot(g, nodeAttrs = nAttrs, attrs = attrs)
```

The scripts are typically run indirectly via `make`, but instructions for
running them individually can be found at the top of each script.

Data files are stored under `analysis/raw`, `analysis/intermediate`, and
`analysis/out`. Most intermediate files are stored in sqlite format so that they
can be easily queried by the viewer app.
